A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain

### 2020-04-20 Environmental-Correlates-of-COVID19-Spain/Environmental-Correlates-of-COVID19-Spain-Title-Page.Rmd
Revised abstract before submission

---
title: A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Fernando A. Lopez
    email: fernando.lopez@upct.es
    affiliation: Universidad Politecnica de Cartagena
  - name: Tatiane Menezes
    email: tatiane.menezes@ufpe.br
    affiliation: Departamento de Economia
  - name: Renata Cavalcanti
    email: renata.vcsantos@gmail.com
    affiliation: Nucleo de Pesquisa
  - name: Maira Galdino da Rocha Pitta
    email: mgrpitta@ufpe.br
    affiliation: Nucleo de Pesquisa
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, 1281 Main St W, Hamilton, ON, L8S 4K1, Canada
  - code: Universidad Politecnica de Cartagena
    address: Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3, 30201, Cartagena, Murcia, Spain
  - code: Departamento de Economia
    address: Departamento de Economia, Universidade Federal de Pernambuco, Av dos Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil  
  - code: Nucleo de Pesquisa
    address: Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av. Prof. Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP 50670-901, Brazil  
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Spreading with astonishing speed, the novel SARS-CoV2 has swept the globe, causing enormous stress to health systems and prompting social distance guidelines and mandates to arrest its progress. While there is encouraging evidence that early public health interventions have slowed the spread of the virus, this has come at a high cost as the global economy is brought to its knees. How and when to ease restrictions to movement hinges in part on the question whether SARS-CoV2 will display seasonality associated with variations in temperature, humidity, and hours of sunshine. In this research, we address this question by means of a spatial analysis of the incidence of COVID-19 in the provinces in Spain. Use of a spatial Seemingly Unrelated Regressions (SUR) approach allows us to model the incidence of reported cases of the disease per 100,000 population, as a function of temperature and humidity, while controlling for GDP per capita, population density, percentage of older adults in the population, and presence of mass transit systems. An interesting aspect of the spatial SUR approach is that it models incidence as a contagion process. Our results indicate that incidence of the disease is lower at higher temperatures and higher levels of humidity, although coefficients for this variable are significant only in some equations. Sunshine, in contrast, displays a positive association with incidence of the disease. Our control variables also yield interesting insights. Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.    
journal: "Geographical Analysis"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---

>>>>

---
title: A spatio-temporal analysis of the environmental correlates of COVID-19 incidence in Spain
author:
  - name: Antonio Paez
    email: paezha@mcmaster.ca
    affiliation: McMaster University
    footnote: 1
  - name: Fernando A. Lopez
    email: fernando.lopez@upct.es
    affiliation: Universidad Politecnica de Cartagena
  - name: Tatiane Menezes
    email: tatiane.menezes@ufpe.br
    affiliation: Departamento de Economia
  - name: Renata Cavalcanti
    email: renata.vcsantos@gmail.com
    affiliation: Nucleo de Pesquisa
  - name: Maira Galdino da Rocha Pitta
    email: mgrpitta@ufpe.br
    affiliation: Nucleo de Pesquisa
address:
  - code: McMaster University
    address: School of Geography and Earth Sciences, McMaster University, 1281 Main St W, Hamilton, ON, L8S 4K1, Canada
  - code: Universidad Politecnica de Cartagena
    address: Departamento de Metodos Cuantitativos, Ciencias Juridicas, y Lenguas Modernas, Universidad Politecnica de Cartagena, Calle Real Numero 3, 30201, Cartagena, Murcia, Spain
  - code: Departamento de Economia
    address: Departamento de Economia, Universidade Federal de Pernambuco, Av dos Economistas, s/n - Cidade Universitária, Recife - PE, 50670-901, Brasil  
  - code: Nucleo de Pesquisa
    address: Núcleo de Pesquisa em Inovação Terapêutica NUPIT / UFPE, Av. Prof. Moraes Rego, 1235 - Cidade Universitária, Recife, PE, CEP 50670-901, Brazil  
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: |
  Early in 2020 the novel SARS-CoV2 swept the globe, disrupting health systems and the economy. Public health interventions have slowed the spread of the virus, albeit at a high cost. How and when to ease restrictions to movement hinges in part on whether SARS-CoV2 will display seasonality due to variations in temperature, humidity, and hours of sunshine. Here, we address this by means of a spatio-temporal analysis of the incidence of COVID-19 in Spain. Use of spatial Seemingly Unrelated Regressions (SUR) allows us to model the incidence of reported cases of the disease per 100,000 population as an interregional contagion process, in addition to a function of temperature, humidity, and sunshine. In the analysis we also control for GDP per capita, percentage of older adults in the population, population density, and presence of mass transit systems. The results support the hypothesis that incidence of the disease is lower at higher temperatures and higher levels of humidity. Sunshine, in contrast, displays a positive association with incidence of the disease. Our control variables also yield interesting insights. Higher incidence is associated with higher GDP per capita and presence of mass transit systems in the province; in contrast, population density and percentage of older adults display negative associations with incidence of COVID-19.     
journal: "Geographical Analysis"
date: "`r Sys.Date()`"
bibliography: References.bib
#linenumbers: true
numbersections: true
csl: elsevier-harvard.csl
output: rticles::elsevier_article
---
